Active Learning for SAT solver Benchmarking

### 2022-10-07 paper/paper.tex
intro1

Our algorithm is unique in that it is the only one, to the best of our knowledge, that has all of the following features (also refer to Tab.~\ref{tab:requirements}).
Our benchmarking process outputs a \emph{ranking} to compare the new solver against the existing set of solvers as described above.
We optimize the \emph{runtime} that our strategy needs to arrive at its conclusion.
We use problem-instance and runtime \emph{features} $\varphi\!: \mathcal{A} \times \mathcal{I} \rightarrow \mathbb{R}^n$ since such features already have proven to be helpful in classic benchmark-set-selection approaches~\cite{HoosKSS13} as well as in portfolio selection~\cite{CollauttiMMO13,NgokoCT19}.
%JB: I think we can remove this criterion, as it also is satisified by all competitors (and is commented out in table; if we keep criterion, we should also show it in table)
%TF: I still find it worth mentioning. So, I added it back to the table.
Moreover, we select instances \emph{non-randomly} and \emph{incrementally}.
By doing so, we have a greater impact on the properties of the benchmarking approach such as its required runtime.
Sampling instances incrementally means taking runtime information from already done experiments into account when choosing the next.
Rather than solely outputting a binary classification, i.e, the new solver is worse than solver A, we provide a \emph{scoring function} that shows by which margin a solver is worse and how similar it is to existing approaches (\emph{comparability}).
%JB: we should rename the "interpretability" criterion (make it more specific), term "interpretability" can have lots of different meanings and our specific meaning is not very intuitive
%TF: Renamed it to "comparability"
Finally, our strategy incrementally selects only a single benchmark set to compare one new solver $\hat{a}$ with the set of existing solvers $\mathcal{A}$ whose cardinality can be any number (\emph{scalability}).
%JB: for what do we measure cost here?
%TF: made it clearer
We avoid comparing pairs of solvers, which requires a non-constant number of benchmark sets.
%JB: need to discuss this; logarithmic (compare one-vs rest) / log-linear (compare each pair) effort might be sufficient if the binary-search-/recursive-sorting-like procedure used
%TF: "non-constant number of benchmark sets"
Since a new solver idea is rarely best on the first try but requires more or less tweaking, it is desired to obtain a solver ranking fast.
In this way, a new solver idea can be discarded if it performs poorly across the board or may be further tweaked if it shows promising results at least in some cases.

@@

Tab.~\ref{tab:requirements} shows a concise overview of existing approaches with a related goal.
%JB: we can probably cut this part a bit, as we discuss weaknesses of individual approaches later anyway
%TF: it's supposed to be a short overview of what is to come
While traditional benchmark-set-selection~\cite{HoosKSS13} lacks the aspect of runtime optimization, related work within algorithm configuration~\cite{HutterHL11} is concerned with finding better configurations fast.
However, algorithm configuration is not suited for ranking.
There is already existing work employing active learning for SAT solver discrimination~\cite{MatriconAFSH21}.
This approach has some major shortcomings though.
Finally, our approach seeks to provide all stated features.
Note that all approaches already make use of problem-instance features as it showed to be beneficial in past work~\cite{HoosKSS13,CollauttiMMO13,NgokoCT19}.
%JB: as commented above, we should either remove this criterion or show it in the table
%TF: added it back to the table

@@

% In terms of our desired features (also refer to Tab.~\ref{tab:requirements}),

@@

While this approach considers the configuration's \emph{runtime}, is \emph{feature-based}, and \emph{scales linearly} by having a fixed time budget, it is not capable of the other three features listed in Tab.~\ref{tab:requirements}.
First, an algorithm configurator cannot be used to \emph{rank} a new solver since it only looks at promising solvers/configurations rather than the overall average performance.
If the new solver does not belong to the best-performing solvers (new ideas are rarely the best approach right from the start, i.e., they need some tweaking), we do not have any information on how this solver performed.
%JB: "this category" = category of most promising solvers? not 100% clear
%TF: made it clearer
Second, while using a model-based selection strategy to sample configurations, instance sampling is done \emph{randomly}, i.e., without building a model over instances.
In fact, tuning algorithms to a small or non-homogeneous set of instances might lead to over-tuning, i.e., lack of generalization~\cite{eggensperger2019pitfalls}.
And third, configurations within their approach are not \emph{comparable} since algorithm configuration seeks to find solemnly the best performing configuration.
%JB: well, the internal meta-model of the configurator allows to estimate the solver performance for particular configurations (rather than solvers), so the approach is kind of interpretable, though regarding configurations (while for comparison to other solvers, only top-configuration performance should be interesting, so the approach is indeed not interpretable regarding solvers)
%TF: changed

@@

Regarding our desired features of adaptive benchmark-set-selection (Tab.~\ref{tab:requirements}), the aforementioned approach already ticks off the first four features, i.e., it \emph{ranks} solvers, optimizes for \emph{runtime}, is \emph{feature-based}, and uses \emph{incremental non-random} sampling.
Although they provide a confidence metric for their prediction, which makes solvers partially \emph{comparable}, it is not clear how similar two given solvers are or on which instances they behave similarly, as the prediction is solely binary.
Moreover, the major shortcoming lies within it being not \emph{scalable} with the number solvers.
As they compare only pairs of solvers, evaluating a new solver requires a non-constant number of benchmark sets, which in turn requires more runtime.
%JB: maybe rather log-linear than quadratic, see a prior comment in intro
%TF: changed

@@

Turning a regression into a classification problem, discretization is a useful tool.
However, it does not provide much value if the resulting cluster labels are not predictive of the PAR-2 ranking we seek to predict.
In our work, we use a hierarchical-clustering approach with a single-link criterion for $k = 3$.
For the SAT Competition~2022 Anniversary Track instances~\cite{sat2022}, this ranking approach is capable of correctly deciding for almost all pairs of solvers which one is faster (about \SI{97.45}{\%}; $\sigma = \SI{3.68}{\%}$).
Moreover, it decides all solvers correctly if the difference in the cluster-label-ranking scores between two solvers is at least \SI{0.11}{}.
%Refer to Tab.~\ref{tab:discreteranking} for a comparison with the PAR-2 ranking.
The Spearman correlation~\cite{de2016comparing} of cluster-label and PAR-2 ranking is about \SI{0.988} (in comparison to a perfect value of \SI{1}).
This shows that monotonicity and score distances between ranks align quite well between both ranking methods.

@@

To obtain the results in Tab.~\ref{tab:mcc}, we perform cross-validation among all solvers: we assume that all other solver runtimes are known and provide \SI{64}{\%} of ground-truth runtime labels of the target solver for training, use \SI{16}{\%} as a validation set for hyper-parameter tuning, and \SI{20}{\%} as a test set (5-fold cross-validation).
This is repeated for each solver once.
Note that we perform cross-validation on two levels, i.e., among the solvers and the target solver runtimes.
We report the average performances on the test sets.
Since the differences in the MCC-performance distributions of the best and second-best approaches are statistically significant regarding a Wilcoxon signed-rank test with $\alpha = \SI{5}{\%}$, we only use the best approach for all further experiments as a prediction model.

@@

Tab.~\ref{tab:paretotable} also shows the performance of particular hyper-parameter choices for different values of $\delta$.
%JB: some plot for comparison would be better than a truncated table showing a variety hyperparameter configs
Thereby, we observe that for small $\delta$ (runtime more important) our approaches preferably use easy instances to get a good sense of the PAR-2 ranking fast, e.g., row with $\delta = 0.15$.
%JB: the "e.g." part needs more explanation (how can we see that easy instances used? -> fraction runtime vs. fraction instances)
In contrast to that, harder instances are preferred for higher values of $\delta$, i.e., the sampled fraction of runtime is greater than the fraction of instances.
In our opinion, the row with $\delta = 0.85$ represents the best trade-off between runtime and PAR-2 accuracy.
%JB: this is a scientific article, not an opinion piece :-); all deltas represent different trade-offs, so I would not claim to have found a best trade-off (probably depends on user which kind of trade-off is most desired)

@@

\subsubsection{Instance-family importance.}
For all configurations of our approach that sample on average less than \SI{10}{\%} of all 5355 instances, we look at the particular instances and families that are selected by our active-learning approach.
Thereby, the average benchmark set contains about \SI{192.34}{} instances.
Tab.~\ref{tab:familiesa} lists the ten most frequent families that are chosen.
For example, \emph{planning} instances account for on average \SI{19.14}{} of the \SI{192.34}{} benchmark instances.
In contrast to that, Tab.~\ref{tab:familiesb} shows the families that are sampled the most often relative to their occurrences in the underlying dataset.
There are, for example, only nine \emph{circuit-multiplier} instances in the complete dataset.
However, we sample on average \SI{1.59}{} of them.
%JB: there should be some (qualitative) interpretation (based on our results, is the sample interesting/special family-wise or not?) rather than just dropping a table into the paper and mainly explaining its structure

>>>>

% The following belongs to Related Work discussion... already references the table there... will fuse / rework tomorrow
Our algorithm is unique in that it is the only one, to the best of our knowledge, that has all of the following features (also refer to Table~\ref{tab:requirements}).
Our benchmarking process outputs a \emph{ranking} to compare the new solver against the existing set of solvers as described above.
We optimize the \emph{runtime} that our strategy needs to arrive at its conclusion.
We use problem-instance and runtime \emph{features} $\varphi\!: \mathcal{A} \times \mathcal{I} \rightarrow \mathbb{R}^n$ since such features already have proven to be helpful in classic benchmark-set-selection approaches~\cite{HoosKSS13} as well as in portfolio selection~\cite{CollauttiMMO13,NgokoCT19}.
%JB: I think we can remove this criterion, as it also is satisified by all competitors (and is commented out in table; if we keep criterion, we should also show it in table)
%TF: I still find it worth mentioning. So, I added it back to the table.
Moreover, we select instances \emph{non-randomly} and \emph{incrementally}.
By doing so, we have a greater impact on the properties of the benchmarking approach such as its required runtime.
Sampling instances incrementally means taking runtime information from already done experiments into account when choosing the next.
Rather than solely outputting a binary classification, i.e, the new solver is worse than solver A, we provide a \emph{scoring function} that shows by which margin a solver is worse and how similar it is to existing approaches (\emph{comparability}).
%JB: we should rename the "interpretability" criterion (make it more specific), term "interpretability" can have lots of different meanings and our specific meaning is not very intuitive
%TF: Renamed it to "comparability"
Finally, our strategy incrementally selects only a single benchmark set to compare one new solver $\hat{a}$ with the set of existing solvers $\mathcal{A}$ whose cardinality can be any number (\emph{scalability}).
%JB: for what do we measure cost here?
%TF: made it clearer
We avoid comparing pairs of solvers, which requires a non-constant number of benchmark sets.
%JB: need to discuss this; logarithmic (compare one-vs rest) / log-linear (compare each pair) effort might be sufficient if the binary-search-/recursive-sorting-like procedure used
%TF: "non-constant number of benchmark sets"
Since a new solver idea is rarely best on the first try but requires more or less tweaking, it is desired to obtain a solver ranking fast.
In this way, a new solver idea can be discarded if it performs poorly across the board or may be further tweaked if it shows promising results at least in some cases.

@@

Table~\ref{tab:requirements} shows a concise overview of existing approaches with a related goal.
%JB: we can probably cut this part a bit, as we discuss weaknesses of individual approaches later anyway
%TF: it's supposed to be a short overview of what is to come
While traditional benchmark-set-selection~\cite{HoosKSS13} lacks the aspect of runtime optimization, related work within algorithm configuration~\cite{HutterHL11} is concerned with finding better configurations fast.
However, algorithm configuration is not suited for ranking.
There is already existing work employing active learning for SAT solver discrimination~\cite{MatriconAFSH21}.
This approach has some major shortcomings though.
Finally, our approach seeks to provide all stated features.
Note that all approaches already make use of problem-instance features as it showed to be beneficial in past work~\cite{HoosKSS13,CollauttiMMO13,NgokoCT19}.
%JB: as commented above, we should either remove this criterion or show it in the table
%TF: added it back to the table

@@

% In terms of our desired features (also refer to Table~\ref{tab:requirements}),

@@

While this approach considers the configuration's \emph{runtime}, is \emph{feature-based}, and \emph{scales linearly} by having a fixed time budget, it is not capable of the other three features listed in Table~\ref{tab:requirements}.
First, an algorithm configurator cannot be used to \emph{rank} a new solver since it only looks at promising solvers/configurations rather than the overall average performance.
If the new solver does not belong to the best-performing solvers (new ideas are rarely the best approach right from the start, i.e., they need some tweaking), we do not have any information on how this solver performed.
%JB: "this category" = category of most promising solvers? not 100% clear
%TF: made it clearer
Second, while using a model-based selection strategy to sample configurations, instance sampling is done \emph{randomly}, i.e., without building a model over instances.
In fact, tuning algorithms to a small or non-homogeneous set of instances might lead to over-tuning, i.e., lack of generalization~\cite{eggensperger2019pitfalls}.
And third, configurations within their approach are not \emph{comparable} since algorithm configuration seeks to find solemnly the best performing configuration.
%JB: well, the internal meta-model of the configurator allows to estimate the solver performance for particular configurations (rather than solvers), so the approach is kind of interpretable, though regarding configurations (while for comparison to other solvers, only top-configuration performance should be interesting, so the approach is indeed not interpretable regarding solvers)
%TF: changed

@@

Regarding our desired features of adaptive benchmark-set-selection (Table~\ref{tab:requirements}), the aforementioned approach already ticks off the first four features, i.e., it \emph{ranks} solvers, optimizes for \emph{runtime}, is \emph{feature-based}, and uses \emph{incremental non-random} sampling.
Although they provide a confidence metric for their prediction, which makes solvers partially \emph{comparable}, it is not clear how similar two given solvers are or on which instances they behave similarly, as the prediction is solely binary.
Moreover, the major shortcoming lies within it being not \emph{scalable} with the number solvers.
As they compare only pairs of solvers, evaluating a new solver requires a non-constant number of benchmark sets, which in turn requires more runtime.
%JB: maybe rather log-linear than quadratic, see a prior comment in intro
%TF: changed

@@

Turning a regression into a classification problem, discretization is a useful tool.
However, it does not provide much value if the resulting cluster labels are not predictive of the PAR-2 ranking we seek to predict.
In our work, we use a hierarchical-clustering approach with a single-link criterion for $k = 3$.
For the SAT Competition~2022 Anniversary Track instances~\cite{sat2022}, this ranking approach is capable of correctly deciding for almost all pairs of solvers which one is faster (about \SI{97.45}{\%}; $\sigma = \SI{3.68}{\%}$).
Moreover, it decides all solvers correctly if the difference in the cluster-label-ranking scores between two solvers is at least \SI{0.11}{}.
%Refer to Table~\ref{tab:discreteranking} for a comparison with the PAR-2 ranking.
The Spearman correlation~\cite{de2016comparing} of cluster-label and PAR-2 ranking is about \SI{0.988} (in comparison to a perfect value of \SI{1}).
This shows that monotonicity and score distances between ranks align quite well between both ranking methods.

@@

To obtain the results in Table~\ref{tab:mcc}, we perform cross-validation among all solvers: we assume that all other solver runtimes are known and provide \SI{64}{\%} of ground-truth runtime labels of the target solver for training, use \SI{16}{\%} as a validation set for hyper-parameter tuning, and \SI{20}{\%} as a test set (5-fold cross-validation).
This is repeated for each solver once.
Note that we perform cross-validation on two levels, i.e., among the solvers and the target solver runtimes.
We report the average performances on the test sets.
Since the differences in the MCC-performance distributions of the best and second-best approaches are statistically significant regarding a Wilcoxon signed-rank test with $\alpha = \SI{5}{\%}$, we only use the best approach for all further experiments as a prediction model.

@@

Table~\ref{tab:paretotable} also shows the performance of particular hyper-parameter choices for different values of $\delta$.
%JB: some plot for comparison would be better than a truncated table showing a variety hyperparameter configs
Thereby, we observe that for small $\delta$ (runtime more important) our approaches preferably use easy instances to get a good sense of the PAR-2 ranking fast, e.g., row with $\delta = 0.15$.
%JB: the "e.g." part needs more explanation (how can we see that easy instances used? -> fraction runtime vs. fraction instances)
In contrast to that, harder instances are preferred for higher values of $\delta$, i.e., the sampled fraction of runtime is greater than the fraction of instances.
In our opinion, the row with $\delta = 0.85$ represents the best trade-off between runtime and PAR-2 accuracy.
%JB: this is a scientific article, not an opinion piece :-); all deltas represent different trade-offs, so I would not claim to have found a best trade-off (probably depends on user which kind of trade-off is most desired)

@@

\subsubsection{Instance-family importance.}
For all configurations of our approach that sample on average less than \SI{10}{\%} of all 5355 instances, we look at the particular instances and families that are selected by our active-learning approach.
Thereby, the average benchmark set contains about \SI{192.34}{} instances.
Table~\ref{tab:familiesa} lists the ten most frequent families that are chosen.
For example, \emph{planning} instances account for on average \SI{19.14}{} of the \SI{192.34}{} benchmark instances.
In contrast to that, Table~\ref{tab:familiesb} shows the families that are sampled the most often relative to their occurrences in the underlying dataset.
There are, for example, only nine \emph{circuit-multiplier} instances in the complete dataset.
However, we sample on average \SI{1.59}{} of them.
%JB: there should be some (qualitative) interpretation (based on our results, is the sample interesting/special family-wise or not?) rather than just dropping a table into the paper and mainly explaining its structure
